Measuring Directionality in Double-Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-08-15 Direction_v7.tex
Read through one more time, and add ref for Manfred.  --Lindley

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double-beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different technique and application than the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz}.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. FIG.~\ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match the KamLAND
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene
(1,2,4-trimethylbenzene) and 1.52~g/l PPO (2,5-diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}. 
Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window. For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This indicates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

The results presented in this paper rely on a simple vertex
reconstruction algorithm, commonly known as a ``point
fit''\cite{SuperKalgo}. It assumes that all of the scintillation and
Cherenkov light is emitted from a single point in space-time
$(x_0,y_0,z_0,t_0)$. In actuality, the light is emitted along an
extended, multi-scattered electron track. However, at the energies
discussed in this paper, the extent of this track is small (a few cm)
compared to the scale of the detector (R=6.5 meters) and thus typical
photon transit distances.

>>>>

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double-beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different technique and application than the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz,dcDirection}.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. FIG.~\ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene
(1,2,4-trimethylbenzene) and 1.52~g/l PPO (2,5-diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}. 
Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window. For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This demonstrates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

The results presented in this paper rely on a simple vertex
reconstruction algorithm, commonly known as a ``point
fit''\cite{SuperKalgo}. It assumes that all of the scintillation and
Cherenkov light is emitted from a single point in space-time
$(x_0,y_0,z_0,t_0)$. In actuality, the light is emitted along an
extended, multi-scattered electron track. However, at the energies
discussed in this paper, the extent of this track is small (a few cm)
compared to the scale of the detector (R=6.5 meters) and therefore the typical
photon transit distances.
